---
title: "Gravity equation tutorial"
author: "Neil Rankin"
date: "March 22, 2017"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(reshape2)
library(stringr)
```

## 1. Introduction

This handout goes through the main steps in empirical trade analysis by using the
gravity model.

The first version of the code was written in Stata (by Malte Ehrich). For most this is likely to be the best choice of statistical software. However, you are free to use whatever package (provided it allows for scripts) you are most comfortable. I will try and add R snippets for those who are R users. [^1]

Sometimes very tiny software-specific steps are ommited for convenience but should not affect your understanding. The full set of steps is present in the Stata do-file.

The tutorial has three parts:

1) Construction of a gravity data set;
2) Estimation of three gravity equations;
3) interpretation of output tables.

The main reference is Bacchetta *et al* (2012).

## 2. Research motivation

**Question:**

How does membership of the World Trade Organisation (WTO) affect trade?

**Hypotheses:**

1. Trade is higher if both countries are members of the WTO.
2. Trade is lower (than when both countries are WTO members) if only one country is a member.

## 3. Construction of a gravity dataset

### Data sources

- Trade flows: [Comtrade-database](https://comtrade.un.org/)
- GDP: [Worldbank](http://data.worldbank.org/indicator/NY.GDP.PCAP.PP.KD), current US-$
- Gravity variables: [CEPII](www.cepii.fr) Distance and other proxies for trade costs, e.g. dummies for colonial history etc.
- WTO membership: [WTO](https://www.wto.org)
- Tariffs: [World Integrated Trade Solution (WITS)](http://wits.worldbank.org/default.aspx) Worldbank


### Loading data (part 1 in do file)
Most raw data are provided in .csv or Excel format.

**Stata example:**

`insheet using "tradeflows.csv", clear delimiter(";") names* `

Label your variables - you'll quickly forget what they mean if you don't 

`label var imports "Imports value in US-$"`

Now, save the data set 

`save "tradeflows.dta", replace`

Do the same for all the remaining data sets (GDP, CEPII,WTOmembership, tariffs,...).

**R example:**
```{r load_data, echo=TRUE}
tradeflows <- read.csv("Data/tradeflows.csv", sep = ";")

```

### Create country-pair combinations (part 2 in do file)

**Table: Mini-example of unbalanced panel**

| Importer     | Exporter | Year | Imports |
| -------------|----------|------|---------|
| South Africa | Germany  | 2000 | 6543    |
| South Africa | Germany  | 2001 | 23434   |
| South Africa | Germany  | 2002 | 665462  |
| South Africa | France   | 1999 | 5321    |
| South Africa | France   | 2002 | 62134   | 



Observations for the country-pairs South Africa-Germany for 1999 and South Africa-France for 2000 and 2001 are missing. We need to replace those with zeros to have a balanced dataset. The idea here is to take into account cases of 0 trade. In this example this is likely to create two challenges. The first is one of memory - this will create a dataset of 1,000,000 lines. Your computer should be able to handle this but size will increase exponentially as the country-pair numbers increase. The second is that if you take the ln(0) this vale will be NA and thus not included in the regression (thus making this exercise moot).

**Stata code** 

`fillin Importer Exporter Year` 

`replace Imports=0 if Imports==.` 

`/*could alternatively make Imports a very small number so that you could ln it*/` 

`replace Imports=0.0001 if Imports==.` 

`save "gravity_temp1.dta", replace`


**R code**
```{r fillin}

vals <- expand.grid(importer = unique(tradeflows$importer),
                    exporter = unique(tradeflows$importer),
                    year = unique(tradeflows$year))
tradeflows1 <- dplyr::left_join(vals,tradeflows)
tradeflows1$imports <- ifelse(is.na(tradeflows1$imports), 0.0001, tradeflows1$imports)
tradeflows1 <- tradeflows1 %>% filter(imports>0) %>% mutate(ln_imports=log(imports))


```

**Table: Mini-example of balanced panel**

| Importer     | Exporter | Year | Imports |
|--------------|----------|------|---------|
| South Africa | Germany  | 1999 |  0      |
| South Africa | Germany  | 2000 | 6543    |
| South Africa | Germany  | 2001 | 23434   |
| South Africa | Germany  | 2002 | 665462  |
| South Africa | France   | 1999 | 5321    |
| South Africa | France   | 2000 | 0       |
| South Africa | France   | 2001 | 0       |
| South Africa | France   | 2002 | 62134   |



### Reshape and merge country-specific data with trade flows

- GDP-data usually comes in wide-format

**Table: GDP in wide-format**

| Country      | Year_1960 | Year_1961 | Year_1962 |
|--------------|-----------|-----------|-----------|
| South Africa | 45646     | 456456    | 4564523   |
| France       | 563456    | 456566    | 45556436  |
| Denmark      | 56546     | 35321     | 96434     |

To merge GDP with trade flow-data, we need to reshape GDP from wide to long format. [^2]

**Stata** 

`reshape long Year, i(countrycode) j(Year)`

**R** 

```{r melt_chunk}

library(readstata13)
GDP <- read.dta13("Data/GDP.dta")

long_GDP <- melt(GDP)
long_GDP <- dplyr::rename(long_GDP, year = variable, GDP = value)
#using stringr package (another package of the tidyverse)
library(stringr)
long_GDP$year <- as.numeric(str_sub(long_GDP$year, 6, 9))



```

**Table: GDP in long-format**

| Country      | Year | GDP      |
|--------------|------|----------|
| South Africa | 1960 | 45646    |
| South Africa | 1961 | 456456   |
| South Africa | 1962 | 4564523  |
| France       | 1960 | 563456   |
| France       | 1961 | 456566   |
| France       | 1962 | 45556436 |
| Denmark      | 1960 | 56546    |
| Denmark      | 1961 | 35321    |
| Denmark      | 1962 | 96434    |


Save the reshaped GDP dataset

**Stata**

`save "GDP_new.dta", replace`

Or if you are using R then you can have multiple objects open at the same time (a **HUGE** benefit) so you don't have to save.

Now, we create GDP for importers and exporter

**Stata**

`use "GDP_new.dta", clear` 

`rename country Exporter` 

`rename GDP GDP_Exporter` 

`save "GDP_Exporter.dta", replace`


`use "GDP_new.dta", clear` 

`rename country Importer` 

`rename GDP GDP_Importer` 

`save "GDP_Importer.dta", replace`


**R**

```{r exp_imp}
#using dplyr here
GDP_exporter <- long_GDP %>% dplyr::rename(GDP_exp = GDP, exporter = countrycode) %>% filter(GDP_exp > 0) %>%mutate(ln_GDP_exp = log(GDP_exp))
#now only keeping the variables we want
GDP_exporter <- GDP_exporter %>% select(exporter, year, ln_GDP_exp)

#importer
GDP_importer <- long_GDP %>% dplyr::rename(GDP_imp = GDP, importer = countrycode) %>% filter(GDP_imp > 0) %>%mutate(ln_GDP_imp = log(GDP_imp))
#now only keeping the variables we want
GDP_importer <- GDP_importer %>% select(importer, year, ln_GDP_imp)

```

Merge trade and GDP datasets.

**Stata**

`use "gravity_temp1.dta", clear`

`sort Exporter Year` 

`merge m:1 Exporter Year using "GDP_exporter.dta"`


We only keep exporter-year pairs for which we have observations in both datasets.


**Stata**

`keep if _merge == 3` 

`drop _merge`


Now, we add GDP-importer.


**Stata**

`sort Importer Year` 

`merge m:1 Importer Year using "GDP_Importer.dta"` 

`keep if _merge == 3` 

`drop _merge` 

`sort Exporter Importer Year` 

`save "gravity_temp2.dta", replace` 


Now we add the standard gravity-type variables (e.g. language, border, etc.).


**Stata**

`use "gravity_temp2.dta", clear` 

`sort Exporter Importer Year` 

`merge m:1 Exporter Importer using "CEPII.dta"` 

`keep if _merge == 3` 

`drop _merge` 

`sort Exporter Importer Year` 

`merge m:1 Exporter Importer using "Religion.dta"` 

`keep if _merge == 1 | _merge == 3` 

`drop _merge` 

`replace Religion = 0 if Religion == .` 

`sort Exporter Importer Year` 

`save "gravity_temp3.dta", replace` 


Before we run our first regression, we need to take logs of imports, GDP, and distance to linearise our model.


**Stata**

`gen ln_Imports = log(Imports)` 

`label var ln_Import "Log of Imports value"` 

`gen ln_GDP_Exporter = log(GDP_Exporter)` 

`label var ln_GDP_Exporter "log of Exporter's GDP"` 

`gen ln_GDP_Importer = log(GDP_Importer)` 

`label var ln_GDP_Importer "log of Importer's GDP"` 

`gen ln_Dist = log(dist)` 

`label var ln_Dist "log of distance"` 

`save "gravity_temp3.dta", replace`


**R** 

That was a lot of work in Stata, let's see what we can do in R.

```{r merging}
data1 <- left_join(tradeflows1, GDP_exporter, by = c("exporter", "year"))
data1 <- left_join(data1, GDP_importer, by = c("importer", "year"))
# do for CEPII, religion etc
cepii <- readstata13::read.dta13("Data/dist_cepii224.dta")
cepii <- cepii %>% rename(importer = country, exporter = partner)

data1 <- left_join(data1, cepii, by = c("exporter", "importer"))

data1$ln_Dist <- log(data1$dist)

```



We run our first regression that we call “Tinbergen”.

**Stata**

`reg ln_Imports ln_GDP_Exporter ln_GDP_Importer ln_Dist` 

`outreg2* creates our tables and needs to be performed after each regression.` 

`outreg2 ln_GDP_Exporter ln_GDP_Importer ln_Dist using Stellenbosch, tex addtext(Year FE, No, Year FE, No) e(N) ctitle("Tinbergen") replace` 


| VARIABLES       | Tinbergen |
|-----------------|-----------|
| ln_GDP_Exporter | 1.092***  |
|                 | (0.00190) |
| ln_GDP_Importer | 0.885***  |
|                 | (0.00184) |
| ln_Dist         | -1.340*** |
|                 | (0.00514) |
| Constant        | -28.42*** |
|                 | (0.0798)  |
| Observations    | 291,859   |
| R-squared       | 0.630     |
| Importer-Year FE|  No       |
| Exporter-Year FE|  No       |
| N               | 291859    |
Standard errors in parentheses
*** p<0.01, ** p<0.05, * p<0.1

**R**

```{r regression, results='asis'}

model1 <- lm(ln_imports ~ ln_GDP_exp + ln_GDP_imp + ln_Dist,  data1, na.action=na.omit)
library(stargazer)
stargazer(model1)
```


Let's add further dummies for trade costs (column (2). 

**Stata**

`reg ln_Imports ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off` 

`outreg2 ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off using Stellenbosch, tex addtext(Year FE, No, Year FE, No) e(N) ctitle("Tinbergen-extended")` 


**R**


```{r regression2, results='asis'}

model2 <- lm(ln_imports ~ ln_GDP_exp + ln_GDP_imp + ln_Dist + colony + contig + comlang_off,  data1, na.action=na.omit)
stargazer(model1, model2)
```


Let's generate our target variables. The dummy “onein” equals one if one of both trading partners is member of the WTO in year t , “bothin” equals one if both are members of the WTO in year t and “nonein” equals one if both are not member of theWTO in year t.

**Stata**

`use "joinWTO.dta", clear` 

`rename country Exporter` 

`rename join join_Exporter` 

`save "joinWTO_Exporter.dta", replace` 

`use "joinWTO.dta", clear` 

`rename country Importer ` 

`rename join join_Importer` 

`save "joinWTO_Importer.dta", replace` 

`use "gravity_temp3.dta", clear` 

`sort Exporter Year` 

`merge m:1 Exporter using "joinWTO_Exporter.dta"` 

`drop _merge` 
`sort Importer Year` 

`merge m:1 Importer using "joinWTO_Importer.dta"` 

`drop _merge` 

`sort Exporter Importer Year` 

`save "gravity_temp4.dta", replace` 

`use "gravity_temp4.dta", clear` 

`foreach var in onein bothin nonein {` 

`gen 'var' = 0 }` 

`replace onein = 1 if (join_Exporter <= Year & join_Importer > Year) |` 

`(join_Importer <= Year & join_Exporter > Year)` 

`label var onein "one of the country pair is member of the WTO"` 

`replace bothin = 1 if (join_Exporter <= Year & join_Importer <= Year)`  

`label var bothin "both countries is member of the WTO"` 

`replace nonein = 1 if (join_Exporter > Year & join_Importer > Year)` 

`label var nonein "none of the country pair is member of the WTO"` 

`save "gravity.dta", replace` 

Lets run the next regression which includes our target variables (column (3),WTO). 

**Stata** 

`reg ln_Imports ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off bothin onein`
`outreg2 ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off bothin onein using Stellenbosch, tex addtext(Year FE, No, Year FE, No) e(N) ctitle("WTO")`

We have learned that one should control for 186 multilateral resistance. Therefore, we need country- and year-dummies.

**Stata**
`use "gravity.dta", clear`

`tab Exporter, gen(Exporter_)`

`tab Importer, gen(Importer_)`

`tab Year, gen(Year_)`

`save "gravity.dta", replace`

How does this affect our estimation results (column (4),MRT)?

`reg ln_Imports ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off bothin onein Exporter_* Importer_* Year_*`
`outreg2 ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off bothin onein using Stellenbosch, tex addtext(Year FE, Yes, Importer FE, Yes, Exporter FE, Yes ) e(N) ctitle("MRT")`

**R example**

```{r dummy_fixed_effects, results='asis'}
#have not created WTO variables in here yet
model3 <- lm(ln_imports ~ ln_GDP_exp + ln_GDP_imp + ln_Dist + colony + contig + comlang_off + as.factor(exporter) + as.factor(importer),  data1, na.action=na.omit)
stargazer(model1, model2, model3)
```

We will now make use of panel data.


$$
\begin{split}
lnX_{ijt} = \beta_{0} + \beta_{1}lnGDP_{it} + \beta_{2}lnGDP_{jt} + \beta_{3}lnDist_{ij} + \beta_{4}Colony_{ij} + \beta_{5}Language_{ij} +  \beta_{6}Contiguity_{ij} \\
+ \beta_{7}onein_{ijt} + \beta_{8}bothin_{ijt} + \beta_{9}nonein_{ijt} \\ 
+ \mu_{i} + \eta_{j} + \nu_{t} + \epsilon_{ijt}
\end{split}
$$

Define the panel structure

**Stata**

`egen pairid = group(Importer Exporter)`

`xtset pairid Year`

Within-group estimation (fixed effects)

**Stata**

`xtreg ln_Imports ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off onein bothin nonein Exporter_* Importer_* Year_*, fe`

`outreg2 ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off bothin onein using Stellenbosch, tex addtext(Year FE, Yes, Importer FE, Yes, Exporter Year, Yes) e(N) ctitle("FE")`

Random effects
`xtreg ln_Imports ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off onein bothin nonein Exporter_* Importer_* Year_*, re`
`outreg2 ln_GDP_Exporter ln_GDP_Importer ln_Dist colony contig comlang_off bothin onein using Stellenbosch, tex addtext(Year FE, Yes, Importer FE, Yes, Exporter Year, Yes) e(N) ctitle("RE")`


**R**

```{r panel_data, results='asis'}
#here it is probably useful to invetsigate the gravity package
```

[^1]:
This document is written in RMarkdown which you should be able to fork from my GitHub account *(neilrankinza)*. There is also a "gravity" package in R (thanks to Ruan Erasmus for pointing this out). I am not really familar with this but it looks like it might be worth investigating if you are using R.

[^2]:
In the data science/R world 'tidy data' is often mentioned. This is a long format where every row is a specific observation (in this case country-year combination) and every column is a variable. There is also an 'extreme' tidy data version used often for packages like ggplot where there is only one column of values and the data is a 'stacked' set of separate varaibles.
